Summary of CA Rice Breeding

### 2017-02-08 output/summary.Rnw
Updated summary with new models and revised figures

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{Sweave}

@@

library(rstan)
library(ggplot2)
library(lme4)
library(rstanarm)

source("../src/prep_model_data.R")
load("h1_fit.Rda")
@

@@

In general, rice yields in CA have been flat over the last 20+ years,
compared to the preceeding period. Yet, annual reports from the CA
Cooperative Rice Research Foundation show that each released variety
has a yield advantage over the previous varieties, typically on the
order of 4--6\%. This raises the question of how to reconcile these
facts. Namely, there are two possibilities:

\begin{enumerate}
\item{There is a mistake in the annual reports, and there is little or no
    yield advantage of newer varieties over old.}
\item{The comparison to old varieties is flawed, allowing an
    appearant yield increase to manifest without an accompanied
    increase in statewide yields.}
\end{enumerate}

\subsection{Easy Facts}

\begin{enumerate}
\item At the year of release (1984), M-202 averaged 9,600 lbs/ac.
\item In 2015, the yield of M-209 was between 9,000-10,000 lbs/ac.
\item Therefore, there does not seem to have been an appreciatable
  increase in yield over this period.
\end{enumerate}

@@

$ yield = \alpha + \beta_{years in trial j}*x_{years in trials j} +
\beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar j}$

Only medium-grained, released varieties were included in the analysis,
reducing the data from \Sexpr{nrow(all_vt_result)} to
\Sexpr{nrow(mod_data)} observations. Included varieties are:
\Sexpr{paste(gsub('^M', 'M-', unique(mod_data[,'id'])), collapse = ', ')}.

@@

<<fig=TRUE, echo=FALSE>>=
plot(mod, regex_pars = "yrs_in_trial")
@ 
\section{H2: Do newer cultivars have a yield advantage?}

@@

In the above model, the estimate for $\beta_{cultivar}$ is
the estimated advantage for cultivar X at year 0 (first year in the
trial) averaged over variability attributable to differences in years,
sites, and the year:site interaction.

% Notes: the expected result here is that there is no difference
% between varieties in year 0. This might be complicated by the fact
% that we have M-202 15 years into its release. We might need to get a
% better way to cross reference these to the true start year.

@@

The raw numbers generally support the conclusion that each new
medium-grain cultivar has an advantage over the previous ones, but
that advantage is not large (Fig. \ref{fig:raw_yield}). Typical
improvements of 5\% are reported by the breeders at the RES, and the
data do not support gains this large.

\begin{figure}

@@

\begin{figure}
<<label=plot2, fig=TRUE, echo=FALSE>>=

@@

However, the question remains how CA yield have not been improving
more after 20 years of breeding effort.

% \begin{figure}

%   <<fig = TRUE, echo=FALSE>>=
%   ## plot(yield_lb ~ year, data = all_vt_result, type = "n")
%   ## mods <- lmList(yield_lb ~ (1|year) | id, data = mod_data)
%   ## pred <- lapply(mods, function(x){
%   ##     data.frame(pred = unique(predict(x)),
%   ##                year = unique(x$model$year))
%   ## })
%   ## sapply(seq_along(pred), function(i) points(x = pred[[i]]$year, y = pred[[i]]$pred,
%   ##                                            pch = i))
%   ## legend("bottomright", legend = vars, pch = 1:5)

% @

% <<echo=FALSE>>=
% ## meds_split <- split(mod_data, paste(all_vt_result[i, "year"], all_vt_result[i,"site"]))
% ## meds_scaled <- lapply(meds_split, function(x){
% ##     m202 <- mean(x$yield_lb[x$id == "M202"])
% ##     x$yield_lb <- x$yield_lb - m202
% ##     return(x)
% ##     })
% ## meds_scaled <- do.call(rbind, meds_scaled)
% ## ggplot(meds_scaled, aes(y = yield_lb, x = year, group = id)) + geom_smooth(aes(color = id), method = "lm") + geom_point()

\begin{enumerate}
  \item Technology improvements over time, including advances in
    nutrient, pest, and water management
  \item Variability over time 
  \item Variability between sites
\end{enumerate}

These factors make it difficult to estimate any one factor separate
from the others. Looking at the raw data, there appears to be a slight
increase in yield over time. From the above, we would expect this due
to technology improvements.

\begin{figure}
<<fig=TRUE, echo=FALSE>>=

by_year <- aggregate(year ~ id, data = all_vt_result, function(x) length(unique(x)))
morethan3 <- by_year$year > 3

i <- all_vt_result$id %in% by_year$id[morethan3]

plot(yield_lb ~ year, mod_data,
     xlab = "Year", ylab = "Yield (lbs/ac)")

abline(lm(yield_lb ~ year, mod_data))

@
\caption{Yield over time for all cultivars during the period 1995 --
  2015. Only named (released) entries are included.}
\end{figure}

\begin{figure}
<<fig=TRUE, echo=FALSE>>=

named <- grepl("^[A-Z]", all_vt_result$id)

named_split <- split(all_vt_result[named,], all_vt_result$id[named])
ans <- lapply(named_split, function(x){
    minYr <- min(x$year, na.rm = TRUE)
    data.frame(yld = x$yield_lb,
               yr = minYr)
})
ans <- do.call(rbind, ans)

plot(yld ~ yr, data = ans,
     xlab = "Year of introduction",
     ylab = "Yield (lbs/ac)")
m1 <- lm(yld ~ yr, data = ans)
abline(m1)
@ 
\caption{The yield of named (released varieties), grouped by year of
  introduction. There is a significant increase over time, but the
  estimate is small (\Sexpr{round(coef(m1)[2], 1)} lbs per year increase).}
\end{figure}

There is evidence of increased yield potential over time, but it is
small, much smaller than the estimated 5\% increase per
release. According to this estimate, there was only a
\Sexpr{round(coef(m1)[2] * 21, 1)} lb increase over
the 21 year period.

\begin{figure}
<<fig=TRUE, echo=FALSE>>=
mod <- lm(yield_lb ~ year, data = all_vt_result)
resids <- residuals(mod)

m202 <- all_vt_result$id == "M202"
par(mfrow = c(1,2))
plot(all_vt_result$yield_lb[m202] ~ all_vt_result$year[m202],
     xlab = "Year",
     ylab = "Yield (raw)",
     main = "M-202") 
abline(lm(all_vt_result$yield_lb[m202] ~ all_vt_result$year[m202]))

plot(resids[m202] ~ all_vt_result$year[m202],
      xlab = "Year",
     ylab = "Yield (de-trended)",
     main = "M-202") 
abline(lm(resids[m202] ~ all_vt_result$year[m202]))
@
\caption{Comparison between the raw yield of cultivar M-202, and the
  yield normalized to remove the annual trends. Amounts less than 0
  are the amount lost compared to using a more modern variety.}
\end{figure}

>>>>

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{Sweave}
\linespread{1.3}

@@

library(rstan)
library(rstanarm)
source("../src/prep_model_data.R")
load("../output/h1_fit.Rda")

@@

In general, rice yields in CA have been flat over the last 20+ years,
compared to the preceding period. Yet, annual reports from the CA
Cooperative Rice Research Foundation show that each released cultivar
has a yield advantage over the previous cultivars, typically on the
order of 4--6\%. This raises the question of how to reconcile these
facts. Namely, there are two possibilities:

\begin{enumerate}
\item{There is a mistake in the annual reports, and there is little or no
    yield advantage of newer cultivars over old.}
\item{The comparison to old cultivars is flawed, allowing an
    apparent yield increase to manifest without an accompanied
    increase in statewide yields.}
\end{enumerate}

\subsection{Motivating Evidence}

\begin{itemize}
\item At the year of release (1984), M-202 averaged 9,600 lbs acre\textsuperscript{-1}.
\item In 2015, the yield of M-209 was between 9,000-10,000 lbs acre\textsuperscript{-1}.
\item Therefore, there does not seem to have been an appreciable
  increase in yield over this period.
\end{itemize}

@@

$$ yield = \alpha + \beta_{years\ in\ trial}* x_{years\ in\ trials} + \beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}$$

Only medium-grained, released cultivars were included in the analysis
and data from the San Joaquin trials, which are not representative of
typical CA rice production (frequent exposure to cold, drill-seeded,
etc.) were excluded. This reduced the data from \Sexpr{nrow(all_vt_result)} to
\Sexpr{nrow(mod_data)} observations. Included cultivars are:
\Sexpr{paste(gsub('^M', 'M-', unique(mod_data[,'id'])), collapse = ', ')}. The
official release years were used for each cultivar for the year of
release (i.e., not the year seed was first commercially available).

@@

\begin{figure}[h]

<<fig=TRUE, echo=FALSE>>=
yrs_in <- -6:34
pp <- posterior_predict(yrs_fit, re.form = NA, newdata = data.frame(yrs_in_trial = yrs_in))
pp <-  pp * 1000 + 9000

ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))

plot(yield_lb ~ yrs_in_trial, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
     xlab = "Year since entry",
     ylab = expression("Yield ( lbs acre"^-1~")"))
lines(x = yrs_in, y = ints[2,], lty = 2) 
polygon(x = c(yrs_in, rev(yrs_in)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
text(20, 3500, bquote(.(round(fixef(yrs_fit)[2] * 1000, 1))~lbs~year^-1))

@ 
\caption{There is a small decrease in yield over the time since a
  cultivar was entered into the trial.}
\label{fig:yrs}
\end{figure}

As seen in Fig. \ref{fig:yrs}, there is a small decrease in yield
over time once a cultivar is released. This decrease is estimated to be
\Sexpr{round(fixef(yrs_fit)[2] * 1000, 1)} lbs
year\textsuperscript{-1}. The oldest cultivar in the trials has lost
an estimated \Sexpr{round(fixef(yrs_fit)[2] * 1000 * 36, 1)} lbs
acre\textsuperscript{-1} since release.

\section{H2: Do newer cultivars have a yield advantage?}

@@

A similar but separate model was used to estimate whether newer
releases had a yield advantage over older ones:

$$ yield = \alpha + \beta_{release\ year}*x_{release\ year} +
\beta_{site} + \beta_{year} + \beta_{site:year} + \beta_{cultivar}$$

Again, the estimate was conditioned on differences between cultivars,
sites, years, and site:years.

@@

\begin{figure}[h]

<<fig=TRUE, echo=FALSE>>=
yrs <- 1980:2015
pp <- posterior_predict(release_fit, re.form = NA, newdata = data.frame(release_yr = yrs))
pp <-  pp * 1000 + 9000

ints <- apply(pp, 2, quantile, p = c(0.05, 0.5, 0.95))

plot(yield_lb ~ release_yr, data = mod_data, pch = 16, col = rgb(0,0,0,0.25),
     xlab = "Year of Release",
     ylab =  expression("Yield ( lbs acre"^-1~")"))
lines(x = yrs, y = ints[2,], lty = 2) 
polygon(x = c(yrs, rev(yrs)), y = c(ints[1,],rev(ints[3,])), col = rgb(0,0,0,0.2), border = NA)
text(1990, 3500, bquote(.(round(fixef(release_fit)[2] * 1000, 1))~lbs~year^-1))
@

\caption{There is an estimated increase in yield over time, but it is
  smaller than the previously estimated 54 lbs
  year\textsuperscript{-1}}
\label{fig:release}
\end{figure}

Contrary to the estimates for decreases once a cultivar is released, there was a
small increase over time (Fig. \ref{fig:release}), though it was small
(\Sexpr{round(fixef(release_fit)[2] * 1000, 1)} lbs
year\textsuperscript{-1}). This suggests that the overall progress on
yield over the period from 1980 -- 2016 is approximately
\Sexpr{round(fixef(release_fit)[2] * 1000 * 36, 1)} lbs acre\textsuperscript{-1}.

\section{Conclusions}

\begin{itemize}
  \item H1: There is evidence of a small decrease in yield once a cultivar is
    released. This complicates comparisons between new entries and
    checks, since the check will decrease over time. This decrease
    will manifest as an apparent larger yield advantage of the new
    entry over the check.
  \item H2: There is evidence of a small increase in yield with each
    new released cultivar. Therefore, it appears that successive
    cultivar releases are not merely recovering to the old benchmark
    but are surpassing it. However, progress is slower than previously
    estimated (\Sexpr{round(fixef(release_fit)[2] * 1000, 1)}
    lbs acre\textsuperscript{-1} year\textsuperscript{-1} increase
    verses a previous estimate of 54 lbs acre\textsuperscript{-1}
    year\textsuperscript{-1}).
\end{itemize}

The practical impact of these results is:

\begin{enumerate}
\item Breeders need to be aware of the potential for the yield of
  checks to decrease over time. Due to this, comparisons to older checks
  might give a false sense of improvement. In order to have greater
  improvements in yield, the target yield improvement for a new cultivar
  needs to be higher to offset the impact of yield decrease once a
  cultivar is released.
\item Breeding is essential to maintaining yield in rice. Yield is not
  stable and without continuous effort, yield potential will decrease
  over time.

\end{enumerate}

\section{Complications}

There is strong collinearity present between the year of release and
the number of years in the trials which is complicating the
simultaneous estimation of these effects. With the current size of the
data set (N = \Sexpr{nrow(mod_data)}), it is impossible to estimate
the decrease over time since release and the improvement in yield over
time with each new release in the same model. This is why each effect was
estimated using a separate regression.

Increasing the size of the data to the full data set (N =
\Sexpr{nrow(all_vt_result)}), allows the simultaneous estimation
and results in similar estimated effect sizes for each of these
mechanisms. However, this necessitates the inclusion of many entries
which have characteristics not similar to conventional cultivars.

\newpage

\section{Supplemental figures}

\begin{figure}[h]

@@

\begin{figure}[h]
<<label=plot2, fig=TRUE, echo=FALSE>>=

@@
